Plots

### 2025-02-25 src/plots.Rmd
Working on plots

---
title: "Metrics"
date: "`r format(Sys.time(), '%B %d, %Y')`"
author: "Richard Wen <richard.wen@unityhealth.to>"
---

@@

## Libraries

```{r}
library(tidyverse)
```

@@

## Settings

```{r}
models_format <- c(
    "_cghr10" = "",
    "_" = " & ",
    "gpt3" = "GPT-3.5",
    "gpt4" = "GPT-4",
    "insilicova" = "InSilicoVA",
    "interva5" = "InterVA-5"
)
stage_format <- c(
    "is_" = "",
    "recon" = "Reconciliation",
    "agreed" = "Agreement",
    "adj" = "Adjudication"
)
```

## Functions

@@

# Calc PCCC
calc_pccc <- function(actual, pred, k = 1, N = NULL) {

@@

    # Calc PCCC
    out <- (C - (k/N)) / (1 - (k/N))
    return(out)
}

@@

    # Calc csmf acc
    out <- 1 - (sum(abs(csmf_true - csmf_pred)) / csmf_max_error)
    return(out)
}

# Bulk calc per model
calc_per_model <- function(actual, pred, func, name = "Metric", ...) {

    # Prep data in long format grouped by model
    out <- pred %>%
        pivot_longer( # to long format
            everything(),
            names_to = "Model",
            values_to = "prediction"
        ) %>% group_by(Model)

    # Calc k if pccc as it is diff for model combos
    if (identical(func, calc_pccc)) {
        out <- out %>%
            summarise(
                !!name := func(
                    actual,
                    prediction,
                    k = str_count(unique(Model), "_"),
                    ...
                )
            )
    } else { # otherwise apply func
        out <- out %>%
            summarise(
                !!name := func(
                    actual,
                    prediction,
                    ...
                )
            )
    }

    # Format display and return
    out <- out %>%
        mutate(Model = str_replace_all(Model, models_format)) %>%
        arrange(across({{name}}, desc))
    return(out)
}

# Bulk calc per model based on age, physician coding stage, and cause of death
calc_by <- function(
        df,
        func,
        name = "Metric",
        by_age = TRUE,
        by_stage = TRUE,
        by_sex = TRUE,
        by_cod = TRUE,
        by_age_range = TRUE,
        by_age_range2 = TRUE,
        ...
    ) {
    out <- list()

    # Calc overall metric
    out[[name]] <- calc_per_model(
        actual = df %>% pull(physician_cghr10),
        pred = df %>% select(ends_with("_cghr10"), -physician_cghr10),
        func = func,
        ...
    ) %>% rename(
        {{ name }} := Metric
    )

    # Calc metric for va coding stage
    if (by_stage) {
        for (stage in c("is_agreed", "is_recon", "is_adj")) {

            # Filter data for stage
            df_filter <- df %>% filter(.[[stage]] == TRUE)
            metric_name = paste0(
                name,
                " ",
                str_replace_all(stage, stage_format)
            )

            # Calc metric for stage
            out[[metric_name]] <- calc_per_model(
                actual = df_filter %>%
                    pull(physician_cghr10),
                pred = df_filter %>%
                    select(ends_with("_cghr10"), -physician_cghr10),
                func = func,
                ...
            ) %>% rename(
                 {{ metric_name }} := Metric
            )
        }
    }

    # Calc pcc for each age group
    if (by_age) {
        for (age_group in c("adult", "child", "neo")) {

            # Filter data for age
            df_filter <- df %>% filter(age == age_group)
            metric_name = paste0(
                name,
                " ",
                str_to_title(age_group)
            )

            # Calc metric for age
            out[[metric_name]] <- calc_per_model(
                actual = df_filter %>%
                    pull(physician_cghr10),
                pred = df_filter %>%
                    select(ends_with("_cghr10"), -physician_cghr10),
                func = func,
                ...
            ) %>% rename(
                 {{ metric_name }} := Metric
            )

            # Calc metric for each stage within age group
            for (stage in c("is_agreed", "is_recon", "is_adj")) {

                # Filter data for stage within age group
                df_filter <- df %>% filter(
                    .[[stage]] == TRUE & age == age_group
                )
                metric_name <- paste0(
                    name,
                    " ",
                    str_to_title(age_group),
                    " ",
                    str_replace_all(stage, stage_format)
                )

                # Calc metric for stage within age group
                out[[metric_name]] <- calc_per_model(
                    actual = df_filter %>%
                        pull(physician_cghr10),
                    pred = df_filter %>%
                        select(ends_with("_cghr10"), -physician_cghr10),
                    func = func,
                    ...
                ) %>% rename(
                     {{ metric_name }} := Metric
                )
            }
        }
    }

    # Calc metric by sex for each age group
    if (by_sex) {

        # Calc metric by cod
        for (a in c("adult", "child", "neo")) {
            for (sx in c("Male", "Female")) {

                # Filter data for cod
                df_filter <- df %>% filter(
                    sex == sx & is_agreed == TRUE & age == a
                )
                metric_name <- sprintf(
                    "%s %s Sex Agree %s", name, str_to_title(a), sx
                )

                # Calc metric for cod
                if (nrow(df_filter) > 0) {
                    out[[metric_name]] <- calc_per_model(
                        actual = df_filter %>%
                            pull(physician_cghr10),
                        pred = df_filter %>%
                            select(ends_with("_cghr10"), -physician_cghr10),
                        func = func,
                        ...
                    ) %>% rename(
                         {{ metric_name }} := Metric
                    )
                }
            }
        }
    }

    # Calc metric by cause of death for each age group
    if (by_cod) {

        # Get unique causes of death
        causes <- df %>%
            select(ends_with("_cghr10")) %>%
            pivot_longer(
                everything(),
                names_to = "column",
                values_to = "cod"
            ) %>%
            distinct(cod) %>%
            filter(!is.na(cod)) %>%
            pull(cod)

        # Calc metric by cod
        for (a in c("adult", "child", "neo")) {
            for (cod in causes) {

                # Filter data for cod
                df_filter <- df %>% filter(
                    physician_cghr10 == cod & is_agreed == TRUE & age == a
                )
                metric_name <- sprintf(
                    "%s %s COD Agree %s", name, str_to_title(a), cod
                )

                # Calc metric for cod
                if (nrow(df_filter) > 0) {
                    out[[metric_name]] <- calc_per_model(
                        actual = df_filter %>%
                            pull(physician_cghr10),
                        pred = df_filter %>%
                            select(ends_with("_cghr10"), -physician_cghr10),
                        func = func,
                        ...
                    ) %>% rename(
                         {{ metric_name }} := Metric
                    )
                }
            }
        }
    }

    # Calc metric by detailed age range
    if (by_age_range) {

        # Get unique age ranges
        age_ranges <- df %>%
            pull(age_range) %>%
            unique

        # Calc metric by age range
        for (a in c("adult", "child", "neo")) {
            for (ar in age_ranges) {

                # Filter for age range
                df_filter <- df %>% filter(
                    age_range == ar & is_agreed == TRUE & age == a
                )
                metric_name <- sprintf(
                    "%s %s Age Agree %s", name, str_to_title(a), ar
                )

                # Calc metric for cod
                if (nrow(df_filter) > 0) {
                    out[[metric_name]] <- calc_per_model(
                        actual = df_filter %>%
                            pull(physician_cghr10),
                        pred = df_filter %>%
                            select(ends_with("_cghr10"), -physician_cghr10),
                        func = func,
                        ...
                    ) %>% rename(
                         {{ metric_name }} := Metric
                    )
                }
            }
        }
    }

    # Calc metric by general age range
    if (by_age_range2) {

        # Get unique age ranges
        age_ranges <- df %>%
            pull(age_range2) %>%
            unique

        # Calc metric by age range
        for (a in c("adult", "child", "neo")) {
            for (ar in age_ranges) {

                # Filter for age range
                df_filter <- df %>% filter(
                    age_range2 == ar & is_agreed == TRUE & age == a
                )
                metric_name <- sprintf(
                    "%s %s Age2 Agree %s", name, str_to_title(a), ar
                )

                # Calc metric for cod
                if (nrow(df_filter) > 0) {
                    out[[metric_name]] <- calc_per_model(
                        actual = df_filter %>%
                            pull(physician_cghr10),
                        pred = df_filter %>%
                            select(ends_with("_cghr10"), -physician_cghr10),
                        func = func,
                        ...
                    ) %>% rename(
                         {{ metric_name }} := Metric
                    )
                }
            }
        }
    }

    # Combine all pccc metrics
    out <- reduce(out,
        function(x, y) left_join(x, y, by = "Model")
    ) %>%
        arrange(across({{ name }}, desc))
    return(out)
}
```

## Data

### Raw Data

@@

```{r}
raw_df <- read_csv("../data/healsl_rd1to2_cod_v1.csv")
```

```{r, echo=FALSE}
cat("Cases: ", nrow(raw_df), "\n")
print(head(raw_df))
```

### Clean Labels

Clean `age_range` labels to include only values and unit of measure in titlecase.

```{r}
df <- raw_df %>% mutate(
    "age_range" = str_to_title(str_replace(
        age_range,
        "\\s*\\(.*\\)",
        ""
    )),
    "age_range" = if_else(
        !str_detect(age_range, "Year|Years|Day|Days|month|Months|week|Weeks"),
        paste0(age_range, " Years"),
        age_range
    ),
    "age_range2" = str_to_title(str_replace(
        age_range2,
        "\\s*\\(.*\\)",
        ""
    )),
    "age_range2" = if_else(
        !str_detect(age_range2, "Year|Years|Day|Days|month|Months|week|Weeks"),
        paste0(age_range2, " Years"),
        age_range2
    )
)
df
```

### Physician Cases

@@

```{r}
df <- df %>%
    filter(!is.na(physician_cghr10) & !is.null(physician_cghr10))
```

```{r, echo=FALSE}
cat("Physician Coded Cases: ", nrow(df))
print(head(df))
```

### Combined Models

@@

```{r}
df <- df %>%
    mutate( # Combine models
        "gpt3_insilicova_cghr10" = if_else(
            gpt3_cghr10 == physician_cghr10 | 
                insilicova_cghr10 == physician_cghr10,
            physician_cghr10,
            gpt3_cghr10
        ),
        "gpt4_insilicova_cghr10" = if_else(
            gpt4_cghr10 == physician_cghr10 | 
                insilicova_cghr10 == physician_cghr10,
            physician_cghr10,
            gpt4_cghr10
        ),
        "gpt3_interva5_cghr10" = if_else(
            gpt3_cghr10 == physician_cghr10 | 
                interva5_cghr10 == physician_cghr10,
            physician_cghr10,
            gpt3_cghr10
        ),
        "gpt4_interva5_cghr10" = if_else(
            gpt4_cghr10 == physician_cghr10 | 
                interva5_cghr10 == physician_cghr10,
            physician_cghr10,
            gpt4_cghr10
        ),
        "gpt3_insilicova_interva5_cghr10" = if_else(
            gpt3_cghr10 == physician_cghr10 | 
                insilicova_cghr10 == physician_cghr10 |
                interva5_cghr10 == physician_cghr10,
            physician_cghr10,
            gpt3_cghr10
        ),
        "gpt4_insilicova_interva5_cghr10" = if_else(
            gpt4_cghr10 == physician_cghr10 | 
                insilicova_cghr10 == physician_cghr10 |
                interva5_cghr10 == physician_cghr10,
            physician_cghr10,
            gpt4_cghr10
        )
    )
```

```{r, echo=FALSE}
print(head(df %>% select(
    gpt3_insilicova_cghr10,
    gpt4_insilicova_cghr10,
    gpt3_interva5_cghr10,
    gpt4_interva5_cghr10,
    gpt3_insilicova_interva5_cghr10,
    gpt4_insilicova_interva5_cghr10
)))
```

### Prepared Data

@@

# Physician info
raw_physicians <- raw_df$physician_cghr10
cat(paste0(
    "Cases (",
    length(unique(na.omit(raw_physicians))), " CODs): ",
    length(raw_physicians),
    "\n"
))

# Physician agreed cases
raw_physicians_agree <- raw_df %>%
    filter(is_agreed == TRUE) %>%
    pull(physician_cghr10)
cat(paste0(
    "Agreed Cases (",
    length(unique(na.omit(raw_physicians_agree))), " CODs): ",
    length(raw_physicians_agree),
    "\n"
))

# Models cases
raw_models <- raw_df %>% select(ends_with("_cghr10"), -physician_cghr10)
for (mcol in colnames(raw_models)) {
    m <- raw_models[[mcol]]
    cat(paste0(
        str_replace_all(mcol, models_format), " Predicted Cases (", 
        length(unique(na.omit(m))), " CODs): ",
        length(m) - sum(is.na(m) | is.null(m)),
        "\n"
    ))
}

@@

## Metrics

Calculate metrics from model outputs compared to physician codes.

```{r}
metrics <- list()
```

### PCCC

Calculate Partial Chance Corrected Concordance (PCCC) to evaluate individual performance for each model by age and physician coding stage.

@@

# Get num of unique causes
ncauses <- df %>%
    select(ends_with("_cghr10")) %>%
    pivot_longer(
        everything(),
        names_to = "column",
        values_to = "cod"
    ) %>%
    distinct(cod) %>%
    filter(!is.na(cod)) %>%
    pull(cod) %>%
    length

# Calc pccc for age, stage, and cod
metrics$pccc <- calc_by(df, calc_pccc, "PCCC", N = ncauses)
metrics$pccc
```

@@

```{r}
metrics$csmf_acc <- calc_by(
    df,
    calc_csmf_acc,
    "CSMF Accuracy",
    by_cod = FALSE
)
metrics$csmf_acc
```

>>>>

---
title: "Plots"
date: "`r format(Sys.time(), '%B %d, %Y')`"
author: "Richard Wen <richard.wen@unityhealth.to>"
---

@@

# Libraries

```{r}
library(tidyverse)
library(dplyr)
library(ggrepel)
```

@@

# Functions

@@

# Calc PCCC
calc_pccc <- function(actual, pred, k = 1, N = NULL, assume_zero = T) {

@@

    # Calc PCCC
    out <- (C - (k/N)) / (1 - (k/N))

    # Scale negative numbers to 0 for easier plotting
    out <- if (assume_zero & out < 0) 0 else out
    return(out)
}

@@

    # Calc csmf acc
    out <- 1 - (sum(abs(csmf_true - csmf_pred)) / csmf_max_error)
    return(out)
}
```

# Data

## Raw Data

@@

# Read preds
df_raw <- read_csv("../data/healsl_rd1to2_cod_v1.csv")

# Display preds
cat("Cases: ", nrow(df_raw), "\n")
print(head(df_raw))

```

## Physician Cases

@@

# Filter for phy cases
df <- df_raw %>%
    filter(!is.na(physician_cghr10) & !is.null(physician_cghr10))

# Preview phy cases
cat("Non-coded Cases: ", nrow(df_raw) - nrow(df))
cat("\nPhysician Coded Cases: ", nrow(df))
df

```

## Combined Models

@@

# Calc combine preds
df <- df %>%
    mutate(
        "gpt3_insilicova_cghr10" = if_else(
            gpt3_cghr10 == physician_cghr10 | 
                insilicova_cghr10 == physician_cghr10,
            physician_cghr10,
            gpt3_cghr10
        ),
        "gpt4_insilicova_cghr10" = if_else(
            gpt4_cghr10 == physician_cghr10 | 
                insilicova_cghr10 == physician_cghr10,
            physician_cghr10,
            gpt4_cghr10
        ),
        "gpt3_interva5_cghr10" = if_else(
            gpt3_cghr10 == physician_cghr10 | 
                interva5_cghr10 == physician_cghr10,
            physician_cghr10,
            gpt3_cghr10
        ),
        "gpt4_interva5_cghr10" = if_else(
            gpt4_cghr10 == physician_cghr10 | 
                interva5_cghr10 == physician_cghr10,
            physician_cghr10,
            gpt4_cghr10
        ),
        "gpt3_insilicova_interva5_cghr10" = if_else(
            gpt3_cghr10 == physician_cghr10 | 
                insilicova_cghr10 == physician_cghr10 |
                interva5_cghr10 == physician_cghr10,
            physician_cghr10,
            gpt3_cghr10
        ),
        "gpt4_insilicova_interva5_cghr10" = if_else(
            gpt4_cghr10 == physician_cghr10 | 
                insilicova_cghr10 == physician_cghr10 |
                interva5_cghr10 == physician_cghr10,
            physician_cghr10,
            gpt4_cghr10
        )
    )

# Display combine preds
print(head(df %>% select(
    gpt3_insilicova_cghr10,
    gpt4_insilicova_cghr10,
    gpt3_interva5_cghr10,
    gpt4_interva5_cghr10,
    gpt3_insilicova_interva5_cghr10,
    gpt4_insilicova_interva5_cghr10
)))

```

## Inspect Data

@@

# Physician info
raw_physicians <- df_raw$physician_cghr10
cat(paste0(
    "Cases (",
    length(unique(na.omit(raw_physicians))), " CODs): ",
    length(raw_physicians),
    "\n"
))

# Physician agreed cases
raw_physicians_agree <- df_raw %>%
    filter(is_agreed == TRUE) %>%
    pull(physician_cghr10)
cat(paste0(
    "Agreed Cases (",
    length(unique(na.omit(raw_physicians_agree))), " CODs): ",
    length(raw_physicians_agree),
    "\n"
))

# Models cases
raw_models <- df_raw %>% select(ends_with("_cghr10"), -physician_cghr10)
for (mcol in colnames(raw_models)) {
    m <- raw_models[[mcol]]
    cat(paste0(
        str_replace_all(mcol, models_format), " Predicted Cases (", 
        length(unique(na.omit(m))), " CODs): ",
        length(m) - sum(is.na(m) | is.null(m)),
        "\n"
    ))
}

@@

# Preparation

Prepare data for plotting.

## Agreed Cases

Filter for agreed cases only.

@@

# Agreed cases only
df <- df %>% filter(is_agreed == TRUE)

```

## COD Case Count

Count the number and percentage of cases for each CGHR-10 COD.

# Count cases per cod
df <- df %>%
    left_join( # count cases and add as column
        df %>%
            group_by(age, physician_cghr10) %>%
            count(name = "physician_cghr10_n") %>%
            group_by(age) %>%
            mutate(
                physician_cghr10_perc = (physician_cghr10_n / sum(physician_cghr10_n)) * 100
            ),
        by = c("age", "physician_cghr10")
    )
df %>% select(rowid, age, physician_cghr10, physician_cghr10_n)
```

# Plots

Generate plots from the predictions.

# Number of causes for each age group
ncod_adult <- 19
ncod_child <- 10
ncod_neo <- 7

# Model full name
gpt3_name <- "GPT-3.5"
gpt4_name <- "GPT-4"
isva_name <- "InSilicoVA"
iva5_name <- "InterVA-5"

# Pivot data into long format for calc
df_long <- df %>%
    filter( # remove low count cod
        physician_cghr10_n >= 10
    ) %>%
    pivot_longer( # long format
        ends_with("_cghr10") & !starts_with("physician_"),
        names_to = "model",
        values_to = "model_cghr10"
    ) %>%
    mutate(
        model = str_replace( # remove cghr10
            model,
            "_cghr10",
            ""
        ),
        model_name = str_replace_all(model, "_", " or ") %>% # use full model names
            str_replace(., "gpt3", gpt3_name) %>%
            str_replace(., "gpt4", gpt4_name) %>%
            str_replace(., "insilicova", isva_name) %>%
            str_replace(., "interva5", iva5_name),
        age_name = case_when( # use full age name
            age == "adult" ~ "Adult",
            age == "child" ~ "Child",
            age == "neo" ~ "Neonate",
            .default = NA
        ),
        cod = physician_cghr10,
        cod_name = paste0(
            cod,
            "\n(n=", physician_cghr10_n, ", ", round(physician_cghr10_perc), "%)"
        ),
        ncod = case_when( # num of cods for each age group
            age == "adult" ~ ncod_adult,
            age == "child" ~ ncod_child,
            age == "neo" ~ ncod_neo,
            .default = NA
        )
    ) %>%
    filter( # remove combine models
        !str_detect(model_name, " or ")
    )

# Calc num of records for each age
n_all <- df %>% group_by(age) %>% count
n_adult <- n_all %>% filter(age == "adult") %>% pull(n)

# Preview
df_long

```

## PCCC by COD

Age, stage, cod.

# Calc pccc
df_plot <- df_long %>%
    group_by(age_name, cod_name, model_name, .drop = F) %>%
    summarize( # calc pccc
        pccc = calc_pccc(cod, model_cghr10, N = unique(ncod)[1])
    ) %>%
    mutate( # add pccc to model name
        model_label = paste0(model_name, " (", round(pccc, 2), ")")
    ) %>%
    filter(age_name == "Adult")

# Add min and max values for each cause of death
df_ends <- df_plot %>%
    group_by(cod_name) %>%
    summarize(
        min_pccc = min(pccc),
        max_pccc = max(pccc),
        model_min = model_name[which.min(pccc)],
        model_max = model_name[which.max(pccc)]
    ) %>%
    arrange(max_pccc)

# Sort cod by max model pccc
df_plot <- df_plot %>%
    mutate(cod_name = factor(cod_name, levels = df_ends$cod_name))
df_ends <- df_ends %>%
    mutate(cod_name = factor(cod_name, levels = df_ends$cod_name))

# Filter out the mid models
df_mids <- df_plot %>%
    left_join(df_ends, by = "cod_name") %>%
    filter(model_name != model_max & model_name != model_min)

# Plot
ggplot(df_plot, aes(x = pccc, y = cod_name)) +
    geom_point( # mid points
        data = df_mids,
        color = "gray60",
        size = 1
    ) +
    geom_point (# min points
        data = df_ends,
        aes(x = min_pccc),
        size = 1.5
    ) +
    geom_point (# max points
        data = df_ends,
        aes(x = max_pccc),
        size = 1.5
    ) +
    geom_segment( # min max lines
        data = df_ends,
        aes(
            x = min_pccc,
            xend = max_pccc,
            y = cod_name,
            yend = cod_name
        ),
        linewidth = 0.75
    ) +
    geom_text( # min model text
        data = df_ends,
        aes(
            x = min_pccc,
            y = cod_name,
            label = paste0(model_min, " (", round(min_pccc, 2), ")")
        ),
        hjust = 1.2,
        vjust = 0.5,
        size = 3
    ) +
    geom_text( # max model text
        data = df_ends,
        aes(
            x = max_pccc,
            y = cod_name,
            label = paste0(model_max, " (", round(max_pccc, 2), ")")
        ),
        hjust = -0.2,
        vjust = 0.5,
        size = 3
    ) +
    geom_text_repel( # mid model text
        data = df_mids,
        aes(label = model_label),
        color = "gray60",
        max.overlaps = Inf,
        size = 2
    ) +
    scale_x_continuous(expand = expansion(add = 0.2)) +
    labs(
        x = "Partial Chance Corrected Concordance (PCCC)",
        y = paste0(length(unique(df_plot$cod_name)), " Causes of Death for ", n_adult, " Adult Records")
    ) +
    theme_minimal() +
    theme(
        panel.border = element_rect(
            color = "black",
            fill = NA,
            linewidth = 0.5
        ),
        legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    )

```

@@

# Create csmf for each age group
csmf_adult <- list()
csmf_child <- list()
csmf_neo <- list()

```
